---
title: "How We Examined Racial Discrimination in Auto Insurance Prices"
author: Jeff Larson, Julia Angwin, Lauren Kirchner, Surya Mattu for ProPublica and
  Dina Haner, Michael Saccucci, Keith Newsom-Stewart, Andrew Cohen, Martin Romm for
  Consumer Reports
date: "4/5/2017"
output:
  html_document:
    fig_caption: yes
  pdf_document:
    fig_caption: yes
  word_document:
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.width=800/72, fig.height=600/72)
source("./analysis.R")
```

## Introduction

It has long been observed that the same safe driver in a predominantly minority zip code is often charged higher auto insurance premiums than if he or she lived in a non-minority neighborhood. A nationwide study by the [Consumer Federation of America](http://consumerfed.org/wp-content/uploads/2015/11/151118_insuranceinpredominantlyafricanamericancommunities_CFA.pdf) in 2015 found that predominantly African-American neighborhoods pay 70 percent more, on average, for premiums than other areas do. But the insurance industry has long argued that higher premiums in minority neighborhoods are justified by a higher risk of accidents. “Insurance rates are color-blind and solely based on risk,” the [Property Casualty Insurers Association of America](https://www.pciaa.net/pciwebsite/cms/content/viewpage?sitePageId=43349) declared in response to the 2015 study. 

We examined that assertion by studying public and private data on auto insurance risk and rates as well as the algorithms used to set prices in different states. We analyzed aggregate risk data by zip code collected by the insurance commissioners of California, Illinois, Missouri and Texas from  insurers in their states. We compared that information with liability insurance premiums — the sum of bodily injury and property damage quotes — charged by the largest companies by market share in each of the four states.

We found that some insurers were charging statistically significantly higher premiums in predominantly minority zip codes, on average, than in similarly risky non-minority zip codes. The difference in premiums was especially stark in Illinois, where nearly every insurer showed a disparity at every risk level. In California, Texas and Missouri, we found disparities in the riskiest zip codes.

In Illinois, `r nrow(il_ratios %>% filter(ratio > 1.1))` of `r nrow(il_ratios)` companies that we analyzed were charging more than 10 percent higher on average for liability premiums in minority zip codes than in white zip codes on average. We also found that companies from the Allstate, Travelers, Metropolitan, Pekin, and Auto Owners insurance groups were charging 30 percent higher on average in risky minority zip codes than in non-minority zip codes with similar risk. Within the city of Chicago, we found that Berkshire Hathaway (Geico), Progressive, Metropolitan and Farmers showed disparities higher than 10 percent. 

In Missouri, we found `r nrow(mo_ratios %>% filter(ratio > 1.1))` of `r nrow(mo_ratios)`  companies with pricing disparities that didn’t appear to be justified by risk. Berkshire Hathaway, Liberty Mutual and Auto Club Enterprises all had subsidiaries that were charging 30 percent more on average in risky minority neighborhoods compared with similarly risky majority white neighborhoods.

In Texas, we found `r nrow(tx_ratios %>% filter(ratio > 1.1))` of `r nrow(tx_ratios)` companies charging residents of risky minority zip codes more than 10 percent higher on average than residents of similar non-minority zip codes.

And in California, `r nrow(ca_ratios %>% filter(ratio > 1.1))` of `r nrow(ca_ratios)` companies were charging more than 10 percent higher on average in risky minority zip codes than in similar non-minority zip codes. Liberty Mutual was charging residents of these minority zip codes  `r floor(((ca_ratios %>% filter(companies_name=="Liberty Mut Fire Ins Co"))$ratio - 1) * 100)` percent more on average than residents of non-minority zip codes.

## How We Acquired the Data

We obtained data from two commercial data providers, [Quadrant Information Services](https://www.quadinfo.com/) and [S&P Global Inc](http://marketintelligence.spglobal.com/). Quadrant provided us with 30 million premium quotes by zip code from the leading insurance companies across the nation. Those quotes reflected rates for forty-four separate profiles of drivers, varying by such factors as gender, age, and family makeup. We used data from the Census's American Community Survey to make sure that these profiles were a representative slice of the United States.

S&P provided us with rate filing manuals, which describe in detail how each insurer sets rates. Where possible, we used public rate filings to check that the Quadrant premiums were correct.

We also filed public records requests for zip code level risk data with insurance commissioners from the 50 states and Washington D.C. Only California, Illinois, Missouri and Texas provided us with data. The other states all responded that they did not collect such data.

In California, Texas, Missouri and Illinois, we received zip-code level data covering the number of cars insured and payouts by the state’s insurers over the most recent five-year period for which data is available. In Texas and Missouri, this loss data covers 2011 to 2015. California’s most recent data covers 2007 through 2011. In 2011, Illinois changed its data collection requirements, so we only used three years of data, from 2012 through 2014.

In California and Texas, the insurance commissioner provided us with incurred losses, which are claims that were filed in those time periods that have been paid in addition to an estimate of claims that have not been paid. The Illinois and Missouri insurance commissioners collect paid losses -- claims that were paid in the calendar year. 

The California loss data is adjusted for “credibility” in areas with low population. In that state, we used these weighted numbers.

California’s losses are also capped: they only represent payouts of minimum coverage policies. People buying insurance can choose to buy more or less coverage, and that means that their claim payouts can be more or less depending on their policy. California's data is "capped" to a single level of coverage in order to even out these differences. In the other three states, we used the raw loss because we did not have access to capped losses.

In Texas, the risk data only include claims and payouts by top insurers, which the state said accounts for approximately 70 percent of the private passenger automobile market. California says that its data covers 99 percent of the market. In Missouri, the risk data covers more than 98 percent of the market. Illinois did not respond to questions on how much of the market their data covers, but their [regulations](ftp://www.ilga.gov/JCAR/AdminCode/050/05004203sections.html) say that reporting requirements are "are applicable to all direct property and liability business written by insurers licensed by the State of Illinois."

Quadrant's data covered 1,816 zip codes in California, 2,023 zip codes in Texas, 1,413 zip codes in Illinois and 1,009 zip codes in Missouri.

## Previous Work

The question of whether and to what extent disparities in auto insurance premiums between minority neighborhoods and non-minority neighborhoods are due to risk has been widely discussed. Research has been impeded by a lack of public data about risk at a zip code level.

A 1997 [study of auto insurance rates in Texas](http://www.cej-online.org/avail.pdf) found that “drivers in poor and minority communities were disproportionately rejected by standard insurers and forced into the higher cost non-standard” insurance plans that are designed as a last resort for people who can’t otherwise buy insurance.

In 1998, the [National Association of Insurance Commissioners studied a sample of auto and homeowner insurance rates nationwide](http://www.naic.org/documents/prod_serv_special_iaa_pb.pdf) and found that higher minority population of a zip code was correlated with higher premiums, even after controlling for factors related to risk. The report noted that zip code level risk data was not available.

Also in 1998, researchers at the [University of South Carolina](http://scholarcommons.sc.edu/cgi/viewcontent.cgi?article=1000&context=fin_facpub) examined auto insurance rates in Missouri and did not find evidence of racial discrimination. They used zip code level data, including aggregated insurance company losses per zip code, but did not examine individual companies.

In 2007, researchers at the University of California-Los Angeles [examined auto insurance rates](http://onlinelibrary.wiley.com/doi/10.1002/pam.20287/full) in Los Angeles by zip code, and by company. They found that “redlining factors are associated with variations in insurance costs” even after risk factors were taken into account.

## Analysis

Insurance companies calculate premiums based on their prediction of how risky a driver will be. In general, an insurer will start with a base rate and increase or decrease that rate based on an insured’s driving record, demographics, location or other factors.

The types of factors used in these risk algorithms vary widely. Some states allow insurers to use factors aside from driving records such as occupation and credit scores as part of their risk calculations.

In order to control for factors outside of geography, we limited our analysis to a single profile – a 30-year-old female safe driver who is a teacher with a bachelor’s degree and excellent credit, with no accidents or moving violations, and who is purchasing standard coverage with the company for the first time. She drives a 2016 Toyota Camry, and has a fifteen-mile commute, and drives around 13,000 miles a year. She is purchasing a policy for \$100,000 of property damage coverage and \$100,000 to cover medical bills per person up to \$300,000 per accident.

We focused on bodily injury and property damage insurance premiums because most states require drivers to purchase them. Within the insurance industry, the combination of bodily injury and property damage coverage is known as liability insurance.

We compared the premiums charged by each insurer in each zip code with the average risk of liability payouts by all insurers in that zip code. 

We calculated a ratio for each zip code of payouts divided by the number of insured cars garaged in that zip code. In California, this is called “average loss” and is a measure of how expensive it is on average for an insurer to write a policy in that zip code:

$Average\ Loss\ =\ \frac{Dollars\ Paid\ Out for\ Liability\ Claims\ by\ Insurers} {Number\ of\ Cars\ Insured\ by\ Insurers}$ 

To be sure, average loss will not exactly match each individual insurer’s payouts in a given zip code. But since insurers are not required to publicly divulge information about their own losses, average loss is the best approximation of loss available on a geographic basis. 

We then sought to determine whether zip codes with predominantly minority populations were priced differently than zip codes with the same risk and different demographics. 

In California, we used the Department of Insurance’s definition of [“underserved” zip codes](http://www.insurance.ca.gov/0400-news/0200-studies-reports/0800-underserved-comm/). The department identifies 145 “underserved” zip codes that have a minority population greater than 66 percent, a household income below the median, and higher than average proportion of uninsured motorists. 

In Texas, Illinois and Missouri, we used data from the Census Department’s American Community Survey 2014 five-year estimates to identify minority zip codes. In Texas, we followed California’s definition of  defined minority zip codes as having greater than 66 percent minority population to. In Missouri and Illinois, we defined minority zip codes as having greater than 50 percent minority population in order to have a sufficiently large sample size. 

We then created separate statistical models for minority and non-minority zip codes to examine differences in pricing. We fit smoothing splines to each predictor: in this case, Average Loss. Because insurance companies do not publish their underlying modeling procedures, we used these splines for their flexibility. This is how we defined each model: 

$Bodily\ Injury\ Premium = Spline(Average\ Loss) + e$

We limited our analysis to zip codes with average losses ranging from \$25 to \$400, except in Illinois where we limited the window from \$50 to \$300 in an effort to remove outliers. Here are the model predictions:

\pagebreak
```{r}
plot_graph(il, "Illinois")
```
\pagebreak
```{r}
plot_graph(mo, "Missouri")
```
\pagebreak
```{r}
plot_graph(tx, "Texas")
```
\pagebreak
```{r}
plot_graph(ca, "California")
```

\pagebreak

In Illinois, we found that premiums were much higher on average in minority zip codes compared to non-minority zip codes with similar risk. In fact, many insurers’ rates showed little association to risk. 

In the other three states, we saw a similar disparity between minority and non-minority zip codes but only in those with high average losses. We calculated average premiums for minority and non-minority zip codes using predictions from the spline models. This allowed us to calculate a ratio for every company in a state:

$Ratio = \frac{Average\ Minority\ Prediction}{Average\ NonMinority\ Prediction}$

We only calculated these ratios for areas with loss measures greater than \$250, except in Illinois where we lowered this threshold to \$200. We chose these thresholds to ensure that the average loss of the minority zip codes was similar to the average loss of non-minority zip codes. We found that our conclusions were similar when we used linear models instead of splines. 

When we repeated the analysis across all 44 profiles, the results were similar.

The results of this analysis are provided in the table below. Ratios reflecting a disparity of greater than 10 percent are bold and italic. Ratios higher than 5 percent are in italic.
\pagebreak
```{r results='asis'}
results_table <- spread(companies %>% select("Group"=name, "Company Name"=companies_name, state, ratio), state, ratio) %>% 
  arrange(Group)
pandoc.table(results_table, 
             emphasize.italics.cells = which(results_table > 1.05, arr.ind = TRUE),
             emphasize.strong.cells = which(results_table > 1.10, arr.ind = TRUE),
             split.tables=Inf,
             round = 3,
             missing = "-")
```

Higher ratios indicate larger difference between prices in minority neighborhoods to those in similarly risky non-minority neighborhoods, on average. 

The ratios are the starkest in Illinois, which is one of the least regulated states for auto insurance. But some insurers also have high ratios in California.

## An Examination of Affordability

We also used the Quadrant premium data to examine auto insurance affordability by zip code.

We used our 30 million auto insurance quotes to calculate a ratio of annual premiums as a share of median household income. These annual premiums include liability insurance as well as more optional coverage like collision insurance, comprehensive insurance and other premiums that are designed to protect the car owner.

We found that auto insurance quotes were more than twice the share of median household income in minority-majority zip codes (11 percent) on average when compared with households in majority white neighborhoods (5 percent). 

The U.S. Treasury Department has defined auto insurance as affordable if it costs less than 2 percent of household income.

## A Look at California

In California’s highly-regulated insurance market, insurers also must give less weight to location than to driving record. Insurers are allowed [two different factors](https://govt.westlaw.com/calregs/Document/I359674A0D49211DEBC02831C6D6C108E?viewType=FullText&originationContext=documenttoc&transitionType=CategoryPageItem&contextData=(sc.Default)) for location-based risk: one for frequency of claims — number of claims divided by policies written in a year — and another for severity of claims — insurance payouts divided by the number of claims.

California regulations require that company specific data for frequency and severity be "credible", where "credibility" is defined in terms of statistical accuracy and precision as mandated by state [regulations](https://govt.westlaw.com/calregs/Document/I4A7AF760D49211DEBC02831C6D6C108E?viewType=FullText&originationContext=documenttoc&transitionType=CategoryPageItem&contextData=(sc.Default)). These rules dictate that a company has received claims from a minimum of 6000 accidents in a zip code over a certain time period, and that the statistical accuracy of this company-chosen time period meets the state's requirements. If the data does not meet those thresholds, companies must combine their data with the state's.

However in areas where the combined data still isn't credible, companies are able to use a provision of the law that allows insurers to group non-credible zip codes together in order to meet that requirement.

The companies with the highest ratios in our analysis disclosed in their public filings that they use this method. Some companies disclose that they use clustering algorithms such as K-Means to draw these areas of contiguous zip codes. Others don’t disclose their algorithms. 

Often, the companies use these algorithms to group these zip codes with nearby areas that also have few policy-holders, according to insurers’ rate filings. This potentially causes higher risk zip codes in these areas to be assigned the average risk for the region, and in many cases, lower risk than what the state estimates.

## A Look at Chicago

In an effort to control disparate pricing, Illinois passed a law in 1972 requiring each car insurance company to charge the same price for bodily injury insurance throughout Chicago. However, each company’s liability insurance premiums can still vary within the city because drivers are required to buy policies that include both bodily injury and property damage coverage, and there is no similar restriction on the variation of prices for property damage.

Because of Chicago’s unique situation, we analyzed Chicago zip codes separately from the rest of Illinois. We found eight companies that were charging higher prices on average for risky zip codes in predominantly minority neighborhoods compared to similarly risky non-minority zip codes within Chicago (see table above). 

## Limitations

A limitation of this study is that in some cases the spline models we used had relatively low R-squared values (see model fits below). Because these values were so low, overall averages may not be precise estimates.

In order to be confident in our results, we also calculated the medians of the fitted values, the means of the actual values and the medians of the actual values. We compared this extra information to the means we calculated in the study and found that they were similar.

However, these low R-squared values also point to another conclusion. The fact that the two distributions don’t line up suggests that insurance companies’ premiums are weakly related to risk.

Another limitation is that companies may have a different distribution of risk than the state’s aggregate risk numbers. But it is unlikely that those differences would result in a consistent pattern of higher prices for minority neighborhoods. 

While companies do not generally publicly disclose their underlying risk numbers, we found a 2015 rate filing in California from Nationwide that did include its internal loss numbers. This filing showed that Nationwide’s internal numbers were different than the state’s, but it also showed that Nationwide’s internal risk numbers were statistically significantly correlated with the state’s (Pearson: `r round(nw_cor$estimate, 2)`, 95% CI:`r round(nw_cor$conf.int[1], 2)`, `r round(nw_cor$conf.int[2], 2)`).

We ran the same analysis as above on premiums from Nationwide's Allied subsidiary and Nationwide's internal loss and found that the disparity between premiums in risky underserved areas was `r round(nw_ratio$ratio * 100 - 100, 0)` percent higher on average than in similarly risky non-underserved areas.That disparity is slightly greater than the 14 percent we found when comparing Allied premiums to overall state risk data. 


```{r}
ggplot(nw, aes(nw_risk, bi_policy_premium + pd_policy_premium, color=minority)) + 
  geom_point(size=0.5) +
  geom_smooth(method="gam", method.args = list(family = "gaussian"), formula=y~s(x)) + 
  ggtitle("Nationwide Loss") +
  xlab("Internal Risk") +
  ylab("Liability Premium") +
  labs(color="Minority") + 
  theme_minimal()
```
\pagebreak

## Model Diagnostics

```{r}
show_diagnostics <- function(data) {
  pandoc.table(data, 
               emphasize.italics.cells = which(results_table < 0.05, arr.ind = TRUE),
               split.tables=Inf,
               round = 3,
               missing = "-")
}
```
\pagebreak

# Illinois Model

```{r results="asis", fig.cap="Illinois Model"}
show_diagnostics(il_diagnostics)
```
\pagebreak

# Chicago Model

```{r results="asis", fig.cap="Chicago Model"}
show_diagnostics(ch_diagnostics)
```
\pagebreak

# Missouri Model

```{r results="asis"}
show_diagnostics(mo_diagnostics)
```
\pagebreak

# Texas Model

```{r results="asis"}
show_diagnostics(tx_diagnostics)
```
\pagebreak

# California Model

```{r results="asis"}
show_diagnostics(ca_diagnostics)
```