---
title: "08_subsetData"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Preparing your data for ggplot2

This guide will likely have the least amount of code, but it may contain the most difficult concept to grasp.

The simple reason is that ggplot2 asks to have the data you want to plot as the focal point of your dataframe.

To make that clear, let's load in the data and libraries we'll need.

```{r}

library(readr)
library(tidyr)

df <- read_csv("dfCrime.csv")

head(df)

```

The __head__ command shows the first six rows of the dataframe.

df is set up like a typical spreadsheet: Rows contain all the information pertaining to the item in the first column, __Year_Quarter__. Each subsequent column contains only the information described by the column header.

So, _total_RTR_ is the total number of response to resistance incidents. The next three columns - _SOF\_only_, _UOF\_only_ and _Transitions_ - are three categories of RTR incidents. If we add up the totals of these three for each row we would get the number in Total_RTR.

